---
title: "EDA"
output: github_document
date: "2025-10-02"
---


```{r}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

Import the weather data
```{r}
data("weather_df")

weather_df = 
  weather_df |>
  mutate(
    month = lubridate::floor_date(date, unit = "month"))

#lubridate: functions for working with date variables
```


Make plots
```{r}
weather_df |> 
  ggplot(aes(x = prcp)) +
  geom_histogram()
```


Check on extreme values
```{r}
weather_df |> 
  filter(prcp > 1000)
```



Look at data again
```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) +
  geom_point(alpha = .75)
```


## add groups

```{r}
weather_df |> 
  group_by(name, month) #24 months in 3 locations, so 72 groups
```


```{r}
weather_df |> 
  group_by(name) |> 
  summarize(
    n = n() #gives number of observations for each group
  )
```



```{r}
weather_df |> 
  group_by(month) |> 
  summarize(
    n = n()
  )
```



```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    n = n()
  )
```

distinct function

```{r}
weather_df |> 
  group_by(month) |> 
  summarize(
    n = n_distinct(date)
  )
#counts distinct instances, in Jan there are 31 distinct dates
```


You can count directly

```{r}
weather_df |> 
  count(name)
```


## More interesting summaries

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE)
  )
#grouping by name and month: for each of the 24 months for each location what is the mean, median,...
```


This is still a dataframe

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
```

cleaning up table to make more readable
```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) |> 
  knitr::kable(digits = 2) #two decimal points
``` 




## mutate with groups

```{r}
#compare the daily max temperature to the annual average max temperature for each station separately
weather_df |>
  group_by(name) |> 
  mutate(mean_tmax = mean(tmax, na.rm = TRUE),
         center_tmax = tmax - mean_tmax
         ) |> 
  ggplot(aes(x = date, y = center_tmax, color = name)) +
  geom_point()


  
```


Look for cold days

```{r}
#in jan 2021, jan 1st the tmax was 4.4 and that was the 14th coldest day
#adding desc() reverses, so it gives hottest day
weather_df |> 
  group_by(name, month) |> 
  mutate(temp_rank = min_rank(desc(tmin))) |> 
  filter(temp_rank < 2) #gives coldest days



```


What about lags?
compare an observation to it’s previous value
find the day-by-day change in max temperature within each station over the year
```{r}
weather_df |> 
  group_by(name) |> 
  mutate(lagged_tmax = lag(tmax))
```



Use the variables you create
```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_change = tmax - lag(tmax)
  ) |> 
  summarize(
    sd_tmax_change = sd(temp_change, na.rm = TRUE),
    tmax_change_max = max(temp_change, na.rm = TRUE)
  )

#find out what day the max change happened: 
#in mutate: change_rank = min_rank(desc(temp_change)) |>  filter(change_rank < 2)
```



Learning assessment

#revisit pulse
```{r}
pulse_df =
  haven::read_sas("data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) |> 
  mutate(visit = fct_inorder(visit)) #putting visit in order that it appears in the dataset

pulse_df |> 
  group_by(visit) |> 
  summarize(
    mean_bdi = mean(bdi, na.rm = TRUE),
    median_bdi = median(bdi, na.rm = TRUE)
  ) |> 
  knitr::kable(digits = 2)
```








